\chapter{Thermodynamics, a Causal Perspective}
\label{ch-thermo}

For a summary of Thermodynamics, see
\cite{wiki-thermo}.

Modern day books on Thermodynamics
derive its 3 laws from either classical
or quantum statistical mechanics, or using classical or quantum stochastic  equations
(see Chapter \ref{ch-stochastic-diff-eqns}). However,
the 3 laws
were originally derived from causal type
	arguments and experimentation, in much
	the same way that one  derives a bnet
	as a hypothesis which is then tested.
Fig.\ref{fig-texnn-for-thermo} is a bnet for Thermodynamics that
captures some of those causal arguments.
The structural equations for the bnet are printed in blue.

	


\begin{figure}[h!]\centering
	$$\xymatrix{
		&*+[F*:SpringGreen]{\underline{V}}\ar@/_1pc/[dddd]\ar[ddddr]\ar[dl]&*+[F*:SpringGreen]{\underline{\{N_i\}}}\ar[ddddr]\ar[dll]&
		\\
		*+[F*:SpringGreen]{\underline{U}}\ar@/_1pc/[dddrrr]\ar[ddd]\ar[dddr]\ar[dddrr]&&*+[F*:Lavender]{\underline{S}}\ar[dddr]\ar@/_1pc/[dddll]\ar[dddl]\ar[ll]&
		\\
		&&&
		\\
		&*+[F*:Lavender]{\underline{T}}\ar[drr]\ar[dl]\ar[d]&*+[F*:Lavender]{\underline{p}}\ar[dl]\ar[d]&*+[F*:Lavender]{\underline{\{\mu_i\}}}\ar[d]
		\\
		*+[F*:SkyBlue]{\underline{F}}&*+[F*:SkyBlue]{\underline{G}}&*+[F*:SkyBlue]{\underline{H}}&*+[F*:SkyBlue]{\underline{\Phi}}
	}$$
	\caption{Thermodynamics, a causal perspective. Extrinsic variables in green, Intrinsic ones in pink, and Legendre  transforms of $U$ in blue.}
	\label{fig-texnn-for-thermo}
\end{figure}

\begin{subequations}
	
	\begin{equation}\color{blue}
		\Phi = \underbrace{U-TS}_F-\sum_i\mu_iN_i\;\;\text{(Grand Potential)}
		\label{eq-X-fun-thermo}
	\end{equation}
	
	\begin{equation}\color{blue}
		\{\mu_i\} = \pder{U}{\{N_i\}}\;\;\text{(chemical potential for species $i$)}
		\label{eq-m-fun-thermo}
	\end{equation}
	
	\begin{equation}\color{blue}
		\{N_i\} = {\rm prior}\;\;\text{(number of particles of species $i$)}
		\label{eq-N-fun-thermo}
	\end{equation}
	
	\begin{equation}\color{blue}
		F = U-TS\;\;\text{(Helmholtz free energy)}
		\label{eq-F-fun-thermo}
	\end{equation}
	
	\begin{equation}\color{blue}
		G = U+pV-TS\;\;\text{(Gibbs free energy)}
		\label{eq-G-fun-thermo}
	\end{equation}
	
	\begin{equation}\color{blue}
		H = U+pV\;\;\text{(enthalpy)}
		\label{eq-H-fun-thermo}
	\end{equation}
	
	\begin{equation}\color{blue}
		p = -\;\pder{U}{V}\;\;\text{(pressure)}
		\label{eq-p-fun-thermo}
	\end{equation}
	
	\begin{equation}\color{blue}
		S = {\rm prior}\;\;\text{(entropy)}
		\label{eq-S-fun-thermo}
	\end{equation}
	
	\begin{equation}\color{blue}
		T = \pder{U}{S}\;\;\text{(temperature)}
		\label{eq-T-fun-thermo}
	\end{equation}
	
	\begin{equation}\color{blue}
		U = U(S, V, \{N_i\})\;\;\text{(internal energy)}
		\label{eq-U-fun-thermo}
	\end{equation}
	
	\begin{equation}\color{blue}
		V = {\rm prior}\;\;\text{(volume)}
		\label{eq-V-fun-thermo}
	\end{equation}
	
\end{subequations}

\section{Delta bnets for Thermodynamics}

In this book, we refer to bnets whose
nodes are differentials (e.g., $df$) or infinitesimals (e.g., $\Delta f$) as {\bf delta bnets.}

The  math of Thermodynamics
(PDEs  relating 
$U, S, T, V, p, F, G, H$, to name a few) can seem like a dizzying  alphabet
soup to the beginner.
The purpose of this chapter is to
show how
delta bnets can be very helpful in visualizing 
and motivating the math
of Thermodynamics.\footnote{We learned the Thermo math, without the delta bnets, for this chapter 
from an open source book
on Physical Chemistry Ref.\cite{fleming-physical-chemistry}. 
Physical Chemistry  uses a lot of Thermodynamics.} Delta bnets are also
useful for discussing It\^o calculus
(see Chapter \ref{ch-stochastic-diff-eqns})
and the back-propagation algorithm (see
Chapter \ref{ch-backp}). 

We shall refer 
to the bnet shown in Eq.(\ref{eq-2c-graph})
as a 2-collision (2C) graph.
Note 
that the arrows 
of a 2C graph
carry partial derivatives,
and that they represent
a perfect differential  $dz$
in 2 variables $x$ and $y$.


\beq
\begin{array}{ll}
\xymatrix{
dx\ar[dr]|a
&
&
dy\ar[dl]|b
\\
&dz
}
&\quad\quad
\left\{
\begin{array}{l}
a=\partial_{x|y}z=
\left(\pder{z}{x}
\right)_y
\\
b=\partial_{y|x}z=
\left(\pder{z}{y}
\right)_x
\\
dz = a \;dx +  b \;dy
\end{array}
\right.
\end{array}
\label{eq-2c-graph}
\eeq

We can even use delta bnets
with inexact differentials such as a small heat change $\delta Q$. In such cases, the arrows exiting
$\delta Q$ will be dotted
arrows, indicating
that such arrows do not carry a partial
derivative. 

In terms of delta bnets,
the First and Second
Laws of Thermodynamics 
can be expressed as:

\beq
\begin{array}{ll}
\xymatrix{
\delta Q
\ar@{.>}[d]_1   
\ar@{.>}[drr]|{1/T}
&dV
\ar[dl]|{-p}
\\
dU
&&dS\ar[ll]|T
}\quad\quad
\begin{array}{l}
\text{First Law}:\quad
dU = \underbrace{\delta W}_{TdS-pdV} +\delta Q
\\
\text{Second Law:}\quad
dS \geq \frac{\delta Q}{T}
\end{array}
\end{array}
\eeq


\hrule
Tabulated parameters for 
Thermodynamics

\begin{itemize}
\item{\bf Heat Capacities}

\beq
\begin{array}{ll}
\xymatrix{
dT\ar[dr]|{C_V}
&
&
dV\ar[dl]
\\
&dU
}
&\quad\quad
C_V = \partial_{T|V}U
\end{array}
\eeq

\beq
\begin{array}{ll}
\xymatrix{
dT\ar[dr]|{C_p}
&
&
dp\ar[dl]
\\
&dH
}
&\quad\quad
C_p = \partial_{T|p}H
\end{array}
\eeq



\item
{\bf Coefficient of
Thermal Expansion}


\beq
\begin{array}{ll}
\xymatrix{
dT\ar[dr]|{\alp V}
&
&
dp\ar[dl]|{-\kappa_T V}
\\
&dV
}
&\quad\quad
\alp = \partial_{T|p}\ln (V)
\end{array}
\eeq


\item {\bf Isothermal Compressibility}


\beq
\begin{array}{ll}
\xymatrix{
dT\ar[dr]|{\alp V}
&
&
dp\ar[dl]|{-\kappa_T V}
\\
&dV
}\quad\quad
&
\kappa_T=-\partial_{p|T}\ln(V)
\end{array}
\eeq


\item {\bf All of the above in a delta bnet}
\beq
\xymatrix@C=2.5pc{
&dV\ar@/_1pc/[dd]
&
\\
dU\ar[dr]|{C_V}
&
&dS
\\
&dT\ar[uu]|{\alp V}
\ar[dr]|{C_p}
&dp\ar[uul]|{-\kappa_T V}
\\
dF
&dG
&dH
}
\eeq
\end{itemize}

\hrule
First Law (adiabatic $\delta Q=0$ version) and differentials of
Legendre transforms.


Reversible Thermodynamics
defines as follows the 3 Legendre transforms 
$F, G, H$ of 
the energy $U$:

\beq
\begin{array}{ll}
\bcen
\xymatrix@C=3pc{
&V\ar[ld]
\ar@/^1pc/[ddd]
\ar[dddr]
&
\\
U
\ar[dd]
\ar[ddr]
\ar@/^1pc/[ddrr]
&
&S\ar[ll]
\ar@/^1pc/[ddll]
\ar[ddl]
\\
&T\ar[d]\ar[dl]
&p\ar[d]
\ar[dl]
\\
F
&G
&H
}\ecen
\quad\quad
&
\left\{
\begin{array}{l}
U=U(S, V)
\\
F=U-TS
\\
G=U+pV -TS
\\
H=U+pV
\end{array}
\right.
\end{array}
\eeq

Even though 
at first it seems that
the 3 Legendre transforms
$F,G,H$, all depend on more 
than 2 dofs (degrees of freedom), 
once we calculate
their differentials,
we realize that they all depend on 2 dofs only.
For example,
from the definition of 
$F$, you might think it depends on 3 dofs, but 
the following calculation shows that it depends only on 2 dofs.



\beq F=U-TS, \quad dU=TdS-pdV
\eeq
\beq
dF= dU-TdS-SdT=
-pdV-SdT
\eeq
The following equation
gives the differentials of $U$ and of the 3 Legendre transforms $F,G, H$.


\beq
\begin{array}{ll}
\bcen
\xymatrix{
&dV\ar[dl]|{-p}
\ar[lddd]|{-p}
&
\\
dU
&
&dS\ar[ll]|{T}
\ar@/^1.5pc/[dd]|T
\\
&dT
\ar[dl]|{-S}
\ar[d]|{-S}
&dp
\ar[dl]|V
\ar[d]|V
\\
dF
&dG
&dH
}
\ecen
&\quad\quad
\left\{
\begin{array}{l}
dU = TdS - pdV
\\
dF = -pdV -SdT
\\
dG =-S dT + Vdp
\\
dH = Vdp +TdS
\end{array}
\right.
\end{array}
\eeq

\hrule
Maxwell's Relations


\beq
\begin{array}{ll}
\xymatrix{
db
&
&da
\\
dx\ar[dr]|{a}
\ar[u]|{\partial_x b}
&
&dy\ar[dl]|{b}
\ar[u]|{\partial_y a=\partial_x b}
\\
&dz
&
}\quad\quad
&
\boxed{
\partial_{y}
\underbrace{\partial_{x|y}z}_a
=
\partial_{x}
\underbrace{\partial_{y|x}z}_b
}
\end{array}
\eeq



\beq
\begin{array}{ll}
\xymatrix{
-dp
&
&-dS
\\
dT\ar[dr]|{-S}
\ar[u]|{-\partial_{T|V}p}
&
&dV\ar[dl]|{-p}
\ar[u]|{-\partial_{V|T}S}
\\
&dF
&
}\quad\quad
&
\boxed{
\partial_{T|V}p=
\partial_{V|T}S
}
\end{array}
\label{eq-dp-is-ds}
\eeq



\beq
\begin{array}{ll}
\bcen
\xymatrix@C=3pc@R=3pc{
dV\ar@/^1pc/[r]|{a_3}
\ar@/^1pc/[d]|{a_4}
&
dS\ar@/^1pc/[l]|{a_2}
\ar@/^1pc/[d]|{-a_4}
\\
dT\ar@/^1pc/[r]|{a_2}
\ar@/^1pc/[u]|{a_1}
&
dp\ar@/^1pc/[l]|{a_3}
\ar@/^1pc/[u]|{-a_1}
}\quad\quad
\ecen
&
\left\{
\begin{array}{l}
a_1= \partial_{V|S}T=-\partial_{S|V}p
\\
a_2=\partial_{p|S}T=\partial_{S|p}V
\\
a_3=\partial_{V|T}S=\partial_{T|V}p
\\
a_4=-\partial_{p|T}S=
\partial_{T|p}V
\end{array}
\right.
\end{array}
\eeq



\hrule
Rules for computing 
partial derivative relations in 
Thermodynamics (adiabatic, $\delta Q=0$ case).


\begin{itemize}



\item {\bf Type I (Reciprocal Rule)}

 
\begin{claim}
\beq
\begin{array}{ll}
\xymatrix{
dx\ar[dr]|{a}
\ar[rr]|{b_1}
&&
dy\ar[dl]|{b_2, c_2}
\\
&dz\ar@/_1.5pc/[ur]|{c_1}
&
}\quad\quad
\left\{
\begin{array}{l}
a = \partial_{x|y}z
\\
b_1=\partial_{x|z}y,\quad
b_2=\partial_{y|x} z
\\
c_1=\partial_{z|x}y,\quad
c_2=\partial_{y|x} z
\\
\boxed{
a =- b_1b_2,\quad
c_1c_2=1
}
\end{array}
\right.
\end{array}
\eeq
\end{claim}
\proof

\beq
dz= dx
\underbrace{\partial_{x|y}z}_{a}
+dy
\underbrace{\partial_{y|x}z}_{b_2, c_2}
\eeq
\beq
dy=dx
\underbrace{\partial_{x|z}y}_{b_1}
+
dz
\underbrace{\partial_{z|x}y}_{c_1}
\eeq

\beq
dz=[a +b_1b_2]dx + [c_1c_2]dz
\eeq

\qed


\item {\bf Type II (Cyclic Permutation Rule)}.

This rule
is the same as
the Type I rule, but with $dz=0$.

\begin{claim}
\beq
\begin{array}{ll}
\xymatrix{
dx\ar[dr]|{a}
\ar[rr]|{b_1}
&&
dy\ar[dl]|{b_2, c_2}
\\
&dz=0\ar@/_1.5pc/[ur]|{c_1}
&
}\quad\quad
\left\{
\begin{array}{l}
a = \partial_{x|y}z
\\
b_1=\partial_{x|z}y,\quad
b_2=\partial_{y|x} z
\\
\boxed{
a =-b_1b_2
}
\end{array}
\right.
\end{array}
\eeq
\end{claim}
\proof See proof of Type I rule.
\qed

Example:

\beq
\begin{array}{ll}
\xymatrix{
dT\ar[dr]|{\alp V}
\ar[rr]|{\partial_{T|V}p}
&&
dp\ar[dl]|{-\kappa_T V}
\\
&dV=0\ar@/_1.5pc/[ur]
&
}\quad\quad
\boxed{\partial_{T|V}p=
\alp/\kappa_T}
\end{array}
\eeq 

\item {\bf Type III Rule (Parallelogram Rule\footnote{I gave it this name. Ref.\cite{fleming-physical-chemistry}
calls it just Type III rule})}


For this rule,
two $dz$ 
must be considered.
We label each of
the 2 $dz$ nodes with a subscript
listing the arrows
entering the node.
The 2 $dz$ 
are numerically equal
(i.e., $d_{xy}z
=
d_{xw}z$ ).

\begin{claim}
\beq
\begin{array}{ll}
\xymatrix@C=1pc{
dy
\ar@/^1pc/[rrrr]|{s_1}
\ar[dr]|{r}
&
&
dx
\ar[ll]
\ar[rr]|{b_1}
\ar[dl]|{\;a'}
\ar[dr]|{a}
&
&
dw
\ar[dl]|{b_2, s_2}
\ar@/_2pc/[llll]
\\
&
d_{yx}z
\ar@{-}[rr]_{=}
&
&
d_{xw}z
}\quad\quad
&
\left\{
\begin{array}{l}
a'=\partial_{x|y}z
\\
a=\partial_{x|w}z
\\
b_1=\partial_{x|y}w
,\quad
b_2=\partial_{w|x}z
\\
r=\partial_{y|x}z
\\
s_1=\partial_{y|x}w,\quad
s_2=\partial_{w|x}z
\\
\boxed{a'=a+b_1 b_2, \quad r=s_1s_2
}
\end{array}
\right.
\end{array}
\eeq
\end{claim}
\proof

\beq
d_{xy}z = d_{xw}z
\implies
dx
\underbrace{\partial_{x|y}z}_{a'}
+ dy
\underbrace{\partial_{y|x}z}_r
=
dx
\underbrace{\partial_{x|w}z}_{a}
+
dw
\underbrace{\partial_{w|x}z}_{b_2,s_2}
\eeq

\beq
dw= dx
\underbrace{\partial_{x|y}w}_{b_1}+
dy
\underbrace{\partial_{y|x}w}_{s_1}
\eeq

\beq
0=
dx\left[
a'
-a
-
b_1
b_2
\right]
+dy\left[
r
-
s_1s_2
\right]
\eeq

\qed

Example 1:

\beq
\begin{array}{ll}
\xymatrix@C=1pc{
dT
\ar@/^1pc/[rrrr]
\ar[dr]|{r}
&
&
dV
\ar[ll]
\ar[rr]|{\partial_{V|T}S}
\ar[dl]|{\partial_{V|T}U}
\ar[dr]|{-p}
&
&
dS
\ar[dl]|{T}
\ar@/_2pc/[llll]
\\
&
d_{TV}U
\ar@{-}[rr]_{=}
&
&
d_{VS}U
}\quad
&
\boxed{
\begin{array}{l}
\partial_{V|T}U
=-p+ T\partial_{V|T}S
\\
\partial_{V|T}S=
\partial_{T|V}p \quad\text{(Maxwell Rel.)}
\end{array}}
\end{array}
\eeq

Example 2:

\beq
\begin{array}{ll}
\xymatrix@C=1pc{
dT
\ar@/^1pc/[rrrr]
\ar[dr]|{r}
&
&
dp
\ar[ll]
\ar[rr]|{\partial_{p|T}S}
\ar[dl]|{\partial_{p|T}H}
\ar[dr]|{V}
&
&
dS
\ar[dl]|{T}
\ar@/_2pc/[llll]
\\
&
d_{Tp}H
\ar@{-}[rr]_{=}
&
&
d_{pS}H
}\quad
&
\boxed{
\begin{array}{l}
\partial_{p|T}H
=V+ T\partial_{p|T}S
\\
\partial_{p|T}S=
- \partial_{T|p}V
\quad\text{(Maxwell Rel.)}
\end{array}}
\end{array}
\eeq


\end{itemize}



