\documentclass[12pt]{report}
\input{../bayesuvius.sty}
\begin{document}

%\chapter{Renormalization Group for PDEs}
%\label{ch-renorm-group-pde}
%

In this chapter, we will
apply the renormalization group (RG)
theory\footnote{See Chapter \ref{ch-renorm-group-intro}
for an introduction into RG theory.} to solving partial differential equations (PDEs).
In particular, we will analyze the Diffusion Equation (DE) PDE
in one space dimension $x$, with a small discontinuity at $x=0$ 
in the diffusion coefficient $D$.

Let

$\vec{r}=(x,y,z)$


$\nabla = (\partial_{x}, \partial_{y}, \partial_{z})$ and $\nabla^2 = \nabla\cdot\nabla=\partial^2_x + \partial^2_y + \partial^2_z$.

$\eta=$ concentration,
number of particles per unit volume (in real life, the volume 
is 3 dimensional, but for most of this chapter, it will be
assumed 1 dimensional for simplicity.)

$D=\frac{\kappa}{2}=$ diffusion coefficient

Recall the following Physics:
\begin{itemize}
\item {\bf Fick's law} (Flux $\vec{J}$ points from high to low concentrations)

\beq
\vec{J}= -D\nabla \eta
\eeq
\item {\bf Continuity equation} (number of particles in control volume $V$  deceases if flux across surface $\partial V$ of control volume is positive)

\beq
\partial_t \eta = -\nabla\cdot \vec{J}, \quad \partial_t\int_V dV\; \eta =-\int_{\partial V} d\vec{A}\cdot \vec{J}
\eeq

\item
{\bf Diffusion Equation }(DE) (Fick's law + continuity equation)

\beq
\partial_t \eta = \nabla 
\left(
D\nabla \eta
\right)
\eeq
\end{itemize}

If $D$ is constant
\beq
\partial_t\eta=D\nabla^2 \eta
\eeq

For the rest of this chapter, we will assume 1 space dimension so spacetime events will be points $(x,t)\in\RR^2$,
$\nabla^2$ will be replaced by $\partial_x^2$, and $\eta$
will stand for particles per unit length. 
Furthermore, we will use the dimensionless quantity

\beq
\xi=\frac{x}{\sqrt{\kappa t}}
\eeq

\section{DE with constant diffusion coefficient}

In this section, we will solve the DE with 
constant diffusion coefficient $D$. RG theory
is  not useful in this case, but
we present a treatment of this case for comparison with
the case of discontinuous $D$, in which RG theory 
does prove useful.  

Let $P(x; \s^2)$ denote the standard normal distribution
\beq
P(x; \s^2)=\frac{1}{\sqrt{2\pi \s^2}}
\exp\left(\frac{-x^2}{2\s^2}\right)
\eeq


\begin{claim}
The solution to the DE

\beq
\partial_t \eta = \frac{\kappa}{2}\partial_x^2 \eta
\label{eq}
\eeq
 assuming $\kappa$ is constant and no boundaries (i.e. free space, $x\in \RR$), subject to the
initial conditions 

\beq
\eta(x, 0)=\eta_\rvx(x)=
NP(x; \ell^2)
\eeq
is 
\beq
\eta(x,t)=
NP(x;\kappa t + \ell^2)
\label{eq-gen-dol-de-gaussian}\eeq
where $N$ is the total number of particles.
\end{claim}
\proof

You can prove this by showing that the 
alleged solution does satisfy the DE and the initial conditions.

 Note that

\beq
\eta_\rvx(x)\rarrow N\delta(x)\quad \text{as $\ell\rarrow 0$}
\eeq
Also note that if we define

\beq 
N(t)=\int_{-\infty}^{\infty}dx\; \eta(x, t)
\eeq
then

\beq
\partial_t N(t) = \frac{\kappa}{2}\int_{-\infty}^\infty \partial_x^2\eta(x, t)=\frac{\kappa}{2} [\partial_x\eta(x,t)]_{x=-\infty}^{x=\infty}=0
\eeq
so the total number of particles (=
the area under $\eta(x, t)$ curve
for any $t$) doesn't change with time.
\qed

Note that for $t\gg \ell^2/\kappa$ (i.e., 
either after a long time or for short Gaussian widths $\ell$),
Eq.(\ref{eq-gen-dol-de-gaussian}) implies

\beq
\eta(x,t)\sim
NP(x; \kappa t)\quad
\text{as $t\rarrow \infty$ or $\ell\rarrow 0$}
\eeq
We call this an {\bf intermediate times}
approximation, because for truly infinite times, $\eta(x,t)\rarrow \infty$. hence, after intermediate
times, the system forgets about the value
of the initial width $\ell$. As we shall see,
when the diffusion coefficient is discontinuous,
the solution of the DE remembers $\ell$ at intermediate times.

Next, let us see what standard engineering dimensional
analysis has to say about this problem. The dimensions
for quantities associated with this problem are
\beq
\engdim{x}=\engdim{\ell}=L, \quad 
\engdim{t}=T, \quad 
\engdim{\eta}= 1/L,\quad 
\engdim{\kappa }= L^2/T,\quad 
\eeq
From this information,
we can define the following 3 {\bf pi-groups}\footnote{
A pi-group is a
dimensionless quantity obtain by multiplying 
or dividing by fractional powers of multiple variables or constants
that do have dimensions).}

\beq
\Pi =\eta \sqrt{\kappa t} ,\quad
\Pi_1 = \frac{x}{\sqrt{\kappa  t}},\quad
\Pi_2= \frac{\ell}{\sqrt{\kappa  t}}
\eeq
Hence, we expect, if standard  engineering dimensional
analysis is correct (it not  always is. For example,
for a DE with discontinuous diffusion constant $D$, it isn't.)

\beq
\Pi = f(\Pi_1, \Pi_2)\implies \eta = \frac{1}{\sqrt{\kappa t}}f\left(\frac{x}{\sqrt{\kappa t}}, \frac{\ell}{\sqrt{\kappa t}}\right)
\eeq
The general solution Eq.(\ref{eq-gen-dol-de-gaussian})
can indeed be put in this form.

Next, let us consider what RG scaling has to say about this problem. Suppose

\beq
x \to x/b, \quad \ell\rarrow \ell/b,\quad t \to b^{y_t} t,\quad \eta \to b^{y_\eta} \eta,
\eeq
Then

\beq
b^{y_\eta -y_t}\partial_t \eta =
b^{y_\eta+2}\partial^2_x\eta
\eeq
so

\beq
y_t=-2
\eeq
Hence, RG scaling predicts

\beq
\eta(x, t;\ell)= b^{-y_\eta}\eta\left(\frac{x}{b}, \frac{t}{b^2};\frac{\ell}{b}\right)
\eeq
If we choose  $b=\sqrt{\kappa t}$, then
\beq
\eta(x, t; \ell)=(\sqrt{\kappa t})^{-y_\eta}\eta\left(\frac{x}{\sqrt{\kappa t}}, 1;
\frac{\ell}{\sqrt{\kappa t}}
\right)
\eeq
For long but intermediate times, if we assume that memory of 
$\ell$ goes away, then

\beq
\eta(x, t; 0)=(\sqrt{\kappa t})^{-y_\eta}\eta\left(\frac{x}{\sqrt{\kappa t}}, 1;0
\right)
\eeq
Recall that $\engdim{\eta}=1/L$. If we assume there
is no anomalous  contribution to $y_\eta$, then $y_\eta=1$ so
\beq
\eta(x,t)
=
\frac{1}{\sqrt{\kappa t}}
f(\xi), \quad\text{where }\xi = \frac{x}{\sqrt{\kappa t}}
\label{eq-eta-kt-fxi-simple}
\eeq


\begin{claim}
The DE implies that the function $f(\xi)$ 
in Eq.(\ref{eq-eta-kt-fxi-simple}) equals
\beq
f(\xi)= A\exp\left(
\frac{-\xi^2}{2}
\right)= A\exp\left(
\frac{-x^2}{2\kappa t}
\right)
\eeq
\end{claim}
\proof
\beq
\partial_x = \frac{1}{\sqrt{\kappa t}}\partial_\xi
\eeq

\beq
\partial_x \eta
=
\frac{1}{\kappa t}
f'(\xi),\quad
\partial_x^2 \eta
=
\frac{1}{(\kappa t)^{3/2}}
f''(\xi)
\eeq


\beq
\frac{1}{\kappa}\partial_t \eta
=
-\frac{1}{2(\kappa t)^{3/2}}
\left[
f + \xi f'
\right]
\eeq
Plugging these derivatives
into the DE partial differential 
equation (PDE) gives the
following ordinary differential equation (ODE)

\beq
-(f+\xi f')
=
f''
\label{eq-basic-ode-gaussian}
\eeq
It's
easy to check that this ODE
is satisfied by a Gaussian centered at $\xi$ 
with $\s^2=1$.
\qed




\section{DE with discontinuous diffusion coefficient}

In this section, we will find the solution of the DE



\beq
\partial_t \eta=
\partial_x\left(D(x)\partial_x \eta\right)
\eeq
where the diffusion coefficient is discontinuous, 
jumping at the $x=0$ interface

\beq
D(x)
= D_0\cald(x)
\eeq
where 
\beq D_0=\frac{\kappa}{2},\quad
\cald(x)=1 + g\indi_{x>0}=
\begin{cases}
1&\text{if }x<0
\\
1+g&\text{if }x<0
\end{cases}
\eeq

The RG solution to this DE will
be obtained by assuming that it is of the form

\beq \boxed{
\eta(x, t)= \left(\frac{t}{T}\right)^{-\frac12+\alp(g)}f(\xi; g)}
\label{eq-eta-rg-most-general}
\eeq
The parameters $A$ and $T>0$ are arbitrary.
The functions $\alp(g)$ and $f(\xi;g)$
will be calculated separately below.

First, we will calculate $\alp(g)$ to first order in $g$
by solving the DE PDE using the method of Green functions.\footnote{
See Appendix \ref{sec-solving-de-green-fun} 
for an introduction to Green functions.}

Second, we will calculate $f(\xi;g)$ to first
order in $g$ by using RG scaling to eliminate 
one dof (degree of freedom) from the DE PDE, thus 
reducing it to an ODE. (We did something similar 
 when discussing the DE with constant diffusion coefficient $D$.)

Note that Eq.(\ref{eq-eta-rg-most-general})
can be expressed in terms of pi-groups
so it doesn't violate naive EDA
(engineering dimensional analysis). However, 
Eq.(\ref{eq-eta-rg-most-general}) is  a 
far stronger statement
than what naive EDA would suggest.



\subsection{Approximating anomalous dimension $\alpha(g)$}


\beq
\partial_t\eta- \partial_x (D(x)\partial_x\eta)=0
\eeq
so
\beq
\left(\partial_t- D_0\partial_x^2\right)\eta=
\partial_x
\left(gD_0\indi_{x>0}\partial_x \eta\right)
\eeq
Suppose we expand  $\eta(x, t)$ in powers of $g$
\beq
\eta=\eta_0+g \eta_1+ g^2 \eta_2\cdots
\eeq



For the zeroth order in $g$, we get:

\beq
(\partial_t-D_0\partial_x^2)\eta_0=0
\eeq
which is solved by


\beq
\eta_0(x,t)
=NP(x; \kappa t)
\eeq
For the  
first order in $g$, we get:

\beq
(\partial_t-D_0\partial_x^2)\eta_1
=
D_0
\partial_x\left(
\indi_{x>0}\partial_x \eta_0
\right)
\eeq
which is solved via Green functions by

\beq
\eta_1(x,t)
=D_0
\int_0^t dt'
\int_{-\infty}^\infty dx'\;
G(x-x',t-t')
\partial_{x'}\left[
\indi_{x'>0}
\partial_{x'}\eta_0(x',t')
\right]
\eeq
If we integrate the last equation by parts over $x'$,
and assume the integrand vanishes at $x'=\pm \infty$,
we get

\beq
\eta_1(x,t)
=D_0
\int_0^t dt'
\int_{0}^\infty dx'\;
[-\partial_{x'}  G(x-x',t-t')
]
\partial_{x'}\eta_0(x',t')
\eeq
We are solving a DE with a source $S(x,t)$ for free space (no boundaries),
so the appropriate Green function (see
Appendix \ref{sec-solving-de-green-fun})  is a normal
distribution

\beq
G(x-x', t-t')= P(x-x'; \kappa(t-t'))
\eeq
Hence

\beq
-\partial_{x'} G(x-x', t-t')=\frac{(x-x')}{\kappa(t-t')}
P(x-x'; \kappa(t-t'))
\eeq
and 

\beq
\partial_{x'} \eta_0(x',t')
=
-\;\frac{x'}{\kappa t'}N P(x'; \kappa t')
\eeq

Assume most of the contribution to the integral
comes from $(\xi, t')\approx (x,t)$ so approximate

\beq
P(x',; \kappa t')\approx P(x; \kappa t)
\eeq
in the integrand. Hence

\begin{align}
\eta_1(x,t)&= -\;D_0N
P(x; \kappa t) \int_0^t dt'
\int_0^\infty dx'\;\frac{x' (x-x')}{\kappa^2 t' (t-t')}P(x-x';\kappa (t-t'))
\\
&=
-\;\frac{1}{2}N
P(x; \kappa t) \int_0^t \frac{dt'}{t'}
\frac{1}{\sqrt{2\pi}}
\int_{0}^\infty d x'\;\frac{x'(x-x')}{[\kappa(t-t')]^\frac32}
\exp\left(-\;\frac{(x-x')^2}{2\kappa(t-t')}\right)
\\
&=
-\;\frac{1}{2}N
P(x; \kappa t) \int_0^t \frac{dt'}{t'}
\frac{1}{\sqrt{2\pi}}
\underbrace{\int_{0}^\infty d y'\;y'(y-y')
\exp\left(-\;\frac{(y')^2}{2}\right)}_{I(y)}
\label{eq-eta-1-double-int}
\end{align}
where 
\beq
y=\frac{x}{\sqrt{\kappa(t-t')}},\quad
y'=\frac{x'}{\sqrt{\kappa(t-t')}}
\eeq
If we  use the following Gaussian integrals
to evaluate $I(y)$
\beq
\int_0^\infty dy' \; y'\exp\left(-\;\frac{(y')^2}{2}\right)=1,\quad
\int_0^\infty dy' \; (y')^2 \exp\left(-\;\frac{(y')^2}{2}\right)=\sqrt{2}\;\Gamma(3/2)=\sqrt{\frac{\pi}{2}}
\eeq
we get
\beq
I(y)= y-\sqrt{\frac{\pi}{2}}
\eeq
Hence
\beqa
\eta_1(x,t)&=& -\frac{1}{2}
NP(x; \kappa t) \int_0^t \frac{dt'}{t'}
 \left[ \frac{y}{\sqrt{2\pi}}-\frac{1}{2}\right]
\\
&\approx&
\eta_0(x,t)
\frac{1}{4}
\ln t
\\
&=& \eta_0(x,t) a\ln t \quad \text{(Define $a =1/4$)}
\eeqa
Hence

\beq
\eta(x, t) 
= NP(x; \kappa t)[1 + g a\ln t+\calo(g^2)]
\label{eq-eta-secular-terms}
\eeq

Eq.(\ref{eq-eta-secular-terms})
is formally correct but a poor perturbation approach
 because 
$g\ln t$ is the product of the perturbation parameter
$g$ times a logarithm of time which will, for large $t$,
ruin the perturbation series. Such time logarithms 
are called secular terms\footnote{A {\bf secular term} in Physics (and Applied Mathematics) is a term in a perturbative expansion or approximate solution that grows without bound over time. Secular in non-technical contexts means non-religious.
Both meanings are connected if you define the adjective \qt{secular} as 
not bounded by time or by religious canon, extending forever.
The word \qt{secular} comes from the Latin \qt{saeculum} meaning "age" or "long period of time."
} To get rid of the secular terms,
we introduce a time scale $T$, and use
\beq
\ln t
=
\ln\left(\frac{t}{T}\right)+\ln T
\eeq
$T$ might be memory of a parameter in the initial
conditions (for example, $\sqrt{\kappa T} = \ell=$ the
initial width of a Gaussian), or it might
be simply a convenient operating point, close 
to the time we are interested in. Basically,
without $T$, we are expanding $\ln t$ about $t=0$,
but with $T$, we are expanding $\ln t$ at $t=T$.

Note that

\beqa
\eta(x, t) 
&=& NP(x; \kappa t)[1 + g a\ln t+\calo(g^2)]
\\
&\approx &
NP(x; \kappa t) T^{ag} \left(\frac{t}{T}\right)^{ag}
\\
&=&
\frac{N}{\sqrt{2\pi \kappa}}\exp\left(\frac{-x^2}{2\kappa t}\right)
T^{-\frac12+ag} \left(\frac{t}{T}\right)^{-\frac12+ag}
\\
&=&\left(\frac{t}{T}\right)^{-\frac12 +\alpha(g)}f(\xi; g)
\eeqa
where

\beqa
f(\xi;g)&=&\frac{N}{\sqrt{2\pi \kappa}}\exp\left(\frac{-x^2}{2\kappa t}\right)T^{-\frac12+ag} + \calo(g)
\\
&=& A(g) \exp\left(\frac{-x^2}{2\kappa t}\right)
\eeqa
and

\beq
\alp(g)= a g + \calo(g^2)
\eeq
Note also that


\beq
\frac{dA}{d\ln T}
=-\;\frac12 + ag+ \calo(g^2)
\eeq

Note that the secular term $\ln t$
came from $
\int_0^t \frac{dt'}{t'}
$. This mirrors the
loop integrals over momentum $
\int \frac{dq}{q}
$
that we encounter 
when applying RG to Quantum Field Theory (QFT).
The analogy between RG for QFT and for
DE with discontinuous diffusion coefficient $D$,
goes much further. See Table \ref{tab-qft-de-comparison}
for a list of analogous parameters between the two.


% Please add the following required packages to your document preamble:
% \usepackage[table,xcdraw]{xcolor}
% Beamer presentation requires \usepackage{colortbl} instead of \usepackage[table,xcdraw]{xcolor}
\begin{table}[h!]
\begin{tabular}{|l|l|l|}
\hline
 & \cellcolor[HTML]{9AFF99}Quantum Field Theory & \cellcolor[HTML]{9AFF99}DE with discontinuous $D$ \\ \hline
\cellcolor[HTML]{FFFC9E}{\color[HTML]{000000} regulator} & $\eps=4-d$ & no analogous parameter \\ \hline
\cellcolor[HTML]{FFFC9E}{\color[HTML]{000000} renormalization point} & $\mu$ & $T$ \\ \hline
\cellcolor[HTML]{FFFC9E}secular parameter & momentum $q$ in loop integrals & time $t$ \\ \hline
\cellcolor[HTML]{FFFC9E}perturbation parameter & coupling constant $g$ & diffusion coefficient jump $g$ \\ \hline
\end{tabular}
\caption{Comparison between RG for QFT and for
DE with discontinuous diffusion coefficient $D$}
\label{tab-qft-de-comparison}
\end{table}

\subsection{Approximating self-similar function $f(\xi; g)$}
In the previous section we found $\alp(g)$ to first order in $g$ and $f(\xi)$ to zeroth order in $g$. In this section,
we find a better approximation for $f(\xi)$.

We are solving the DE PDE subject to the following
{\bf  transmission (i.e., interface) continuity conditions}:
\begin{itemize}
\item concentration continuity
\beq
\eta(0-, t)=\eta(0+, t)
\eeq
This can be re-expressed in terms of $f(\xi)$ as
\beq
f(0-)=f(0+)
\eeq

\item flux continuity
\beq
D_{-}\partial_x \eta(0-, t) = D_{+}\eta(0+, t)
\eeq
Since 
\beq
\partial_x \eta
=
\frac{1}{\sqrt{\kappa t}}
f'(\xi)
\eeq
this can be re-expressed in terms of $f(\xi)$ as
\beq
\cald_-f'(0-)=\cald_+f'(0+)
\eeq
where


\beq
\cald(\xi)=
\begin{cases}
\cald_-=1& \text{if }\xi<0\\
\cald_+=1+g& \text{if  }\xi>0
\end{cases}
\eeq

\end{itemize}
Note, this is not
a boundary value problem in the usual sense.
There is no boundary at $x=0$. Boundary
value problems normally
have either Dirichlet, von Neumann or mixed boundary 
conditions (see Sec.\ref{sec-solving-de-green-fun}).

If we  rewrite the DE PDE in terms of $f(\xi)$
instead of $\eta(x,t)$ as we
did for the DE with a constant diffusion
coefficient, then we get
the following ODE analogous 
to Eq.(\ref{eq-basic-ode-gaussian})



\beq
-\cald(\xi)(f+\xi f')
=
f''
\eeq
As in the case of the DE 
with a constant diffusion coefficient,
this ODE is satisfied by a Gaussian centered at $\xi=0$
and with $\s^2=D(\xi)$. We must have:

\beq
f(\xi)
=
\begin{cases}
A_-
\exp \left(-\frac{\xi^2}{2\cald_-}\right)
&\text{if }\xi<0
\\
A_+
\exp \left(-\frac{\xi^2}{2\cald_+}\right)
&\text{if }\xi>0
\end{cases}
\eeq
We must also have:

\beq
f(0-)=f(0+)
\eeq
Hence

\beq
A_+=A_- \eqdef A
\eeq
and

\beq
\boxed{f(\xi;g)
=
A
\exp \left(-\frac{\xi^2}{2\cald(\xi)}\right)+\calo(g^2)}
\label{eq-best-f-xi}
\eeq

Since

\beq
f'(\xi)
=
-\frac{\xi}{2\cald(\xi)}
A
\exp \left(-\frac{\xi^2}{2\cald(\xi)}\right)
\eeq
the flux continuity 
is satisfied automatically. It reduces to just  $0=0$. 

We conclude that $f(\xi)$ is just two half Gaussians
joined at their maximum. The left one has width $\cald_-$ and 
the right one has width $\cald_+$.




\section{Appendix: Solving DE via Green Functions}
\label{sec-solving-de-green-fun}

Let
\beq
\call = \partial_t - D \partial^2_x
\eeq
The Diffusion Equation (DE) in one space dimension $x$,
 with a source $S(x,t)$, is 
\beq
\call \eta(x,t) = S(x,t)
\eeq
We will illustrate the use of Green
functions by solving this PDE 
for all $x>0$ and $t>0$, with boundary conditions (BC) 
at $x=0$.

Common types of BCs:

\begin{itemize}
\item {\bf Dirichlet BC}

\beq \eta(0, t)= \eta_\rvt(x)
\eeq

\item
{\bf von Neumann BC}
\beq
\partial_x\eta(x,t)= f(x,t)
\eeq
For example, if there were an impermeable wall at $x=0$,

\beq
\partial_x\eta(x,t)=0
\eeq

\item {\bf mixed BC}

\beq
[a \partial_x\eta + b \eta](x, t)=g(x,t)
\eeq
\end{itemize}





\begin{claim}\label{cl-de-diriclet-green}
The solution to the equation
\beq
\call \eta (x,t) = S(x,t)\quad
\eeq
for $x>0$ and $t>0$
with initial conditions
\beq
\eta(x,0)= \eta_\rvx(x)
\eeq
and Dirichlet boundary conditions:

\beq
\eta(0,t) = \eta_\rvt(t)
\eeq
is

\beqa
\eta(x,t)&=&
\int_0^{t+} dt'\;
\int_0^\infty dx'\;
G(x,t;x',t')
f(x',t')
\\
&&+
\int_0^{\infty} dx'\;
G(x,t;x', 0)
 \eta_{\rvx}(x')
\\
&&-D
\int_0^{t+} dt'\;
[\partial_{x'} G(x,t;x',t')]_{x'=0}
 \eta_\rvt(t')
\eeqa
where $G(x,t;x',t')$ satisfies 
the following equations for $x>0$ and $t>0$


\beq
\call G(x,t;x', t')=\delta(x-x')\delta(t-t')
\eeq
and

\beq
G(x=0,t;x',t')=0
\eeq
\end{claim}
\proof

Define
\beq
\call= \partial_t -D \partial_x^2,\quad
\call G(x,t;x', t')=\delta(x-x')\delta(t-t')
\eeq


\beq
\call^\dagger= -\partial_t -D \partial_x^2,
\quad \call^\dagger G'(x,t;x', t')=\delta(x-x')\delta(t-t')
\eeq



\beq
P_\rvx = \int_0^\infty dx'\; \ket{\rvx=x'}\bra{\rvx=x'}
\eeq


\beq
P_\rvt = \int_0^t dt'\; \ket{\rvt=t'}\bra{\rvt=t'}
\eeq

\beq
\partial_{\rvx} = 
\int_0^\infty dx'\; \ket{\rvx=x'}\partial_{x'}\bra{\rvx=x'}
\eeq

\beq
\partial_{\rvt} = 
\int_0^\infty dt\; \ket{\rvt=t'}\partial_{t'}\bra{\rvt=t'}
\eeq

\beq
\rv{\call}= \partial_\rvt-D\partial^2_\rvx
\eeq

\beq
f(x)\cev{\partial_x} a(x)= a(x) \partial_x f(x)
\eeq
We will use the following suggestive notation
for dual tensor product spaces



\beq \stackt{A}{B}= A\otimes B\eeq

\beq \kett{x}{t}=\ket{x}\otimes\ket{t}\eeq
Written using this notation,  what we want to prove is this
\beq
\boxed{\brat{x}{t}\ket{\eta}=
\brat{x}{t}G
\stackt{P_\rvx}{P_\rvt}\ket{S}
+
\brat{x}{t}G\stackt{P_\rvx}{P_0}\kett{\eta_\rvx}{0}
-D
\brat{x}{t}G\stackt{\cev{\partial}_{\rvx}}{P_\rvt}\kett{0}{\eta_\rvt}}
\eeq

Define $I$, $I_\rvx$ and $I_\rvt$  by
\beq
I = \brat{x}{t}G\rv{\call}
\ket{\eta}-
\bra{\eta}\rv{\call^\dagger}
G'\kett{x}{t}
\eeq

\beq
I_\rvt = 
\brat{x}{t}G
\stackt{1}{\partial_\rvt}
\ket{\eta}-
\bra{\eta}
\stackt{1}{(-\partial_\rvt)}
G'\kett{x}{t}
\eeq

\beq
I_\rvx = 
\brat{x}{t}G\partial_\rvx^2
\ket{\eta}-
\bra{\eta}\partial_\rvx^2
G'\kett{x}{t}
\eeq
Hence 
\beq
\boxed
{I = I_\rvt -D I_\rvx}
\eeq
\hrule
Note that we can simplify the
expression for $I$ as follows:

\beq
\call^\dagger \brat{x}{t}G\kett{x'}{t'
}=\brat{x}{t}\rv{\call}^\dagger G'\kett{x'}{t'}=
\stackt{\delta(x-x')}{\delta(t-t')}
\implies\quad
\rv{\call}^\dagger G' =1
\eeq



\beq
\rv{\call}\ket{\eta}=\ket{S}
\eeq

\beq
\boxed
{I = \brat{x}{t}G\ket{S}-\bra{\eta}\kett{x}{t}}
\eeq
\hrule
Note that we can simplify the
expression for $I_\rvt$ as follows:


\beqa
I_\rvt &=&
\int_0^t dt'
\brat{x}{t}G
\stackt{1}{\ket{t'}\partial_{t'}\bra{t'}}
\ket{\eta}
+
\bra{\eta}
\stackt{1}{\ket{t'}\partial_{t'}\bra{t'}}
G'\kett{x}{t}
\eeqa

\beq
\bra{t}G\ket{t'}= \bra{t'}G'\ket{t}
\eeq

\beq
\partial_{t'} (\ket{t'}\bra{t'})=
\ket{t'}\partial_{t'}\bra{t'}
+\ket{t'}\cev{\partial}_{t'}\bra{t'}
\eeq


\beqa
I_\rvt &=&
\int_0^t dt'
\brat{x}{t}G
\stackt{1}{\partial_{t'} (\ket{t'}\bra{t'})}
\ket{\eta}
\eeqa

\beq
\int_0^t
dt'\;\partial_{t'} (\ket{t'}\bra{t'})
=
\ket{t}\bra{t}-\ket{0}\bra{0}
\eeq


\beq
\boxed{I_\rvt =
\brat{x}{t}G
\stackt{P_\rvx}{\ket{t}\bra{t}-\ket{0}\bra{0}}
\ket{\eta}}
\eeq
\hrule
Note that we can simplify the
expression for $I_\rvx$ as follows:

\beqa
I_\rvx &=& 
\int_0^\infty dx'
\brat{x}{t}G
\stackt{\ket{x'}\partial_{x'}^2\bra{x'}}{1}
\ket{\eta}
-
\bra{\eta}
\stackt{\ket{x'}\partial_{x'}^2\bra{x'}}{1}
G'\kett{x}{t}
\eeqa

\beq
\ket{x'}\partial_{x'}^2\bra{x'}
-\ket{x'}\cev{\partial}_{x'}^2\bra{x'}=
\partial_{x'}(\ket{x'}\partial_{x'}\bra{x'})
-(\ket{x'}\cev{\partial}_{x'}\bra{x'})\cev{\partial}_{x'}
\eeq

\beq
I_\rvx=
\int_0^\infty dx'
\brat{x}{t}G
\stackt{\partial_{x'}(\ket{x'}\partial_{x'}\bra{x'})
-(\ket{x'}\cev{\partial}_{x'}\bra{x'})\cev{\partial}_{x'}}{1}
\ket{\eta}
\eeq

\begin{align}
\int_0^\infty dx'
\partial_{x'}(\ket{x'}\partial_{x'}\bra{x'})
-(\ket{x'}\cev{\partial}_{x'}\bra{x'})\cev{\partial}_{x'}
&=
[\ket{x'}\partial_{x'}\bra{x'}
-
\ket{x'}\cev{\partial}_{x'}\bra{x'}]_0^\infty
\\
&=
-[\ket{x'}\partial_{x'}\bra{x'}
-
\ket{x'}\cev{\partial}_{x'}\bra{x'}]_{x'=0}
\end{align}

\beq
I_\rvx=
\brat{x}{t}G
\stackt{[-\ket{x'}\partial_{x'}\bra{x'}
+
\ket{x'}\cev{\partial}_{x'}\bra{x'}]_{x'=0}}{1}
\ket{\eta}
\eeq


\beq
\boxed{I_\rvx=\brat{x}{t}G
\stackt{[
\ket{x'}\cev{\partial}_{x'}\bra{x'}]_{x'=0}}{P_\rvt}
\ket{\eta}}
\eeq
\qed

Claim \ref{cl-de-diriclet-green} uses a Green function
$G(x,t;x',t')$ with specific properties,
but it does give an explicit expression for $G$
or even prove that such a $G$ exists.
It is easy to check that 
the following $G$ has the desired properties

\beq
G = G^+ - G^-
\eeq
where

\beq
G^\pm(x,t;x',t')
=
\frac{1}{\sqrt{4\pi D(t-t')}}
\exp\left[-\;\frac{(x\pm x')^2}{4D(t-t')}\right]
\eeq

At the boundary
\beq
\left.
\partial_{x'} G
(x,t;x',t')
\right|_{x'=0}
=
\frac{x}{\sqrt{4\pi D (t-t')^3}}
\exp\left(
-\frac{x^2}{4D(t-t')}
\right)
\eeq
This $G$ was found by the method of images: two
free space solutions $G_\pm$ to $\ul{\call}G_\pm=1$
were superposed so as  to satisfy
$G=0$ at the boundary $x=0$. 



\end{document}