
\documentclass[12pt]{report}
\input{bayesuvius.sty}
\begin{document}

\subsection{Do Query Visibility}

Let $G$ be a graph before
amputation
of the arrows entering node $\rvx$,
and let $G_{amp(x)}=\cald_{\rvx=x}G$
be the same graph after amputation
at $\rvx$.
Also let $P_G(\cdot)=P(\cdot|G)$ be
the promise
for graph $G$, and $P_{G_{amp(x)}}(\cdot)=
P(\cdot|G_{amp(x)})$
be that for $G_{amp(x)}$.
In general, the
following is always true,
for bnets
with or  without hidden nodes:\footnote{Note that $P_{G_{amp(x)}}(y|x)\neq P_G(y|x)$
in general.
In fact,
$P_{G_{amp(x)}}(y|x)= P_G(y|x)$
iff there is no confounding,
so $P_{G_{amp(x)}}(y|x)\neq P_G(y|x)$
indicates confounding.
}

\beq
P_{G_{amp(x)}}(y|\cald\rvx=x)=P_{G_{amp(x)}}(y|x)\neq P_G(y|x)
\;
\eeq
However, $P_{G_{amp(x)}}(y|x)$ is
not a valid adjustment formula
for this query
because it's not expressed in terms
of observed TPMs of $G$. It would
require data
observed in a RCT
to 


We define a {\bf do-adjustment formula}
for a {\bf do-query}
$P(y|\cald\rvx=x)$
on a bnet $G$
which we call
the {\bf original promise (OP)}
to be an algebraic expression constructed from the TPMs of non-hidden nodes of $G_{amp(\rvx)}$.

If a do-adjustment formula
exists for a
particular do-query,
then we say the do-query is
 {\bf do-invisible}.
% \footnote{
%To prove that a do-query $P(y|\cald\rvx=x, z)$
%is do-visible
%for a graph $G$,
%just prove that $\rvy\perp\rvx|\rvz$
%in $\call_\rvx G$.
%This is called Rule 2 of Do Calculus,
%but it is easy to understand just from
%the d-separation theorem.
%Info can be transmitted
%between $\rvy$ and $\rvx$ by
%either (1) paths
%in $\cald_\rvx G$
%or (2) paths in $\call_\rvx G$.
%$P(y|\cald\rvx=x, z)=
%P(y|x, z)$
%means the info is being transmitted only
%by (1).
%So the Rule 2 premise is checking
%that no info is being transmitted
%by (2). }
A {\bf do-transfer formula}
is a relationship between 2  do-queries.





Given a bnet $\calg$, let $\cald_\rvx \calg=
\calg_{amp(\rvx)}$
be the bnet $\calg$
after all arrows entering $\rvx$ have been amputated.
$P(y\mid do(\rvx)=x)=P(y\mid \cald \rvx=x)$
corresponds to the
instantiation of $\cald_\rvx \calg$ wherein all nodes other than $\rvx$ and $\rvy$ have been 
summed over.  
Hence, {\bf $P(y\mid \cald x=x)$
is visible} iff
\beq
0=\delta_{hidden}P(y\mid \cald \rvx =x)
\eeq
for all possible $\delta_{hidden}$.
The goal of Pearl's Do Calculus is
two-fold:
\begin{enumerate}
\item 
Decide whether a bnet is visible or not.
Of course, bnets
without hidden
nodes are always visible.

\item If the bnet has hidden
nodes but is visible, find
an equivalent bnet
that gives the same promise
but without hidden nodes. 

\end{enumerate}


Note that Pearl's \qt{identifiability of causal effects} is actually defined
as a partial do-invisibility,
whereas we will always assume 
total identifiabity.

Note also that Pearl's 3 rules of Do Calculus
deal specifically
with do-invisibility
for probabilities conditioned
on one or more do contraints (i.e., constraints of the form $do(\rvx)=x$),
but note that do-invisibility
is an issue that arises for any
probability distribution
with hidden variables, not
just for \qt{doey} ones.



\end{document}