\chapter{Do Calculus proofs}
\label{ch-do-calc-proofs}

In Chapter \ref{ch-do-calc},
we explained Do Calculus,
but referred to this
chapter for proofs
of claims that
use Do Calculus.
In this chapter, we've
aggregated
 all proofs, from
throughout the book,
of claims that use Do Calculus.

Note that even though the 3
rules of Do Calculus
are great for proving
adjustment formulae
for general classes of DAGs,
they are sometimes overkill
for proving
 adjustment formulae
for a single specific DAG.
Indeed, since the
 3 rules of Do Calculus
are a consequence
of the d-separation theorem, it follows that
all adjustment
formulae should be
provable from first principles,
assuming only
the d-separation theorem
and the standard rules of
probability theory.
In this chapter, we present
a technique that allows us to do just that: prove 
adjustment formulae from first principles,
without using the 
3 Rules of Do Calculus.

\section{Bnet Conventions}
In this chapter, we use the
 following conventions for bnet diagrams.

\bnetInstantiations

\hiddenNodes

\selectionGraphs


We often write

\beq
\begin{array}{c}
P(a)\xymatrix {a\ar[r]&\sum b\ar[r]&c}
\quad=\quad
P(a) \xymatrix {a\ar[r]&c}
\\
\sum_b P(c|b)P(b|a)P(a)=P(c,a)
\label{eq-promi-dep}
\end{array}
\eeq
because  the left-hand side summed over $a,c$ yields 1. However,
this equation
is only true if the $P(c,a)$
on the right-hand side
assumes a particular promise, namely
\beq
P(c,b, a)=P(c|b)P(b|a)P(a)
\eeq
An alternative way of saying the same thing
is to say that Eq.(\ref{eq-promi-dep})
 assumes that $P(\cdot)=P_G(\cdot)$ where
$G$ is the bnet $\rva\rarrow\rvb\rarrow\rvc$.
We say that Eq.(\ref{eq-promi-dep})
is {\bf bnet promise dependent}.
\section{Identifiability}

In Statistics, given a probability distribution
$P(x;\theta)$ that depends on a parameter $\theta$, 
we say  {\bf parameter $\theta$ is identifiable } iff

\beq
\forall\; \delta\theta \neq 0,\quad \quad
\delta P\neq 0 
\eeq
Hence, the \qt{slope} $\delta P/\delta \theta \neq 0$
for all $\delta\theta$.\footnote{
This is usually stated equivalently as
$\delta P=0$ implies $\delta\theta=0$.} Equivalently, {\bf parameter $\theta$ is UN-identifiable} 
 iff

\beq\exists \;\delta\theta\in \RR\text{ such that }
\quad\quad \delta P= 0
\eeq

I will say a {\bf a probability distribution $P(\cdot)$ is partially (resp., totally) identifiable} (and the hidden variables
analogous to $\theta$  
are partially (resp., totally) unidentifiable) iff 
\beq \text{partially:}\quad
\exists \text{ changes in the 
hidden variables such that}\quad\quad
\delta P=0
\eeq
\beq
\text{totally:}\quad
\forall \text{ changes in the 
hidden variables,}\quad\quad
\delta P=0
\eeq
Whenever I talk about 
$P(\cdot)$ being
identifiable or unidentifiable
without specifying whether I 
mean partially or totally,
I will be referring to
the total version. 



Let's be more specific about
the $P(\cdot)$ identifiability.


Define an
{\bf infinitesimal variation }
$\delta P$ of the
 probability distribution $P(x)$ by

\beq
\delta P(x)=
\delta x\; \pder{P(x)}{x}
\eeq
\beq
\delta_{P_\rvx} \sum_x f(x) P(x)= \sum_x f(x) \delta P(x)
\eeq

This assumes $P(x)$
is a smooth function of
a continuous  $x$.
If we can prove or
disprove identifiability
for the smooth $P(x)$ case,
that is strong evidence that
it is true in the discrete case too.




Suppose a bnet $\calg$ with nodes 
$\rvx_1, \rvx_2, \ldots, \rvx_N$ has a joint 
distribution (i.e., promise)
given by
\beq
P(x_{1\dots N})
=\prod_{i=1}^N P(x_i\mid pa(x_i))
\eeq
where $pa(x_i)$ is the empty set for root nodes.
Suppose $H\subset \{1, 2, \ldots, N\}$ is the set of indices belonging to the
hidden variables of the bnet.
Then we define $\calg$  
to be {\bf identifiable} iff

\beq
0=\delta_{hidden}P(x_{1\dots N})
=\sum_{h\in H}
\left[\prod_{i\neq h}P(x_i\mid pa(x_i))
\right]\delta
P(x_h\mid pa(x_h))
\eeq  
for all possible 
variations
$\delta P(x_h\mid pa(x_h)$
for all $h\in H$.
Hence, an identifiable bnet
is invariant under changes
in the probability
distributions of its hidden variables. 

Given a bnet $\calg$, let $\cald_\rvx \calg$
be the bnet $\calg$
after all arrows entering $\rvx$ have been amputated.
$P(y\mid do(\rvx)=x)=P(y\mid \cald \rvx=x)$
corresponds to the
instantiation of $\cald_\rvx \calg$ wherein all nodes other than $\rvx$ and $\rvy$ have been 
summed over.  This instantiation
of $\cald_\rvx \calg$ is a
probability distribution represented by
a bnet too, so the definition
of identifiabilty applies to it too.
Hence, {\bf $P(y\mid \cald x=x)$
is identifiable} iff
\beq
0=\delta_{hidden}P(y\mid \cald \rvx =x)
\eeq
for all possible $\delta_{hidden}$.
The goal of Pearl's Do Calculus is
two-fold:
\begin{enumerate}
\item 
Decide whether a bnet is identifiable or not.
Of course, bnets
without hidden
nodes are always identifiable.

\item If the bnet has hidden
nodes but is identifiable, find
an equivalent bnet
that gives the same promise
but without hidden nodes. 

\end{enumerate}

Note that Pearl's \qt{identifiability of causal effects} is actually defined 
as a partial identifiability,
whereas we will always assume 
total identifiabity.

Note also that Pearl's 3 rules of Do Calculus
deal specifically
with identifiability 
for probabilities conditioned
on one or more do contraints (i.e., constraints of the form $do(\rvx)=x$),
but note that identifiability
is an issue that arises for any
probability distribution
with hidden variables, not
just for \qt{doey} ones.



\section{Original Promise and Universal Operations}

So far, we have defined
a promise to be the full
probability distribution of a bnet.
Next, we shall introduce
the concepts of 
{\bf original promise (OP)} and {\bf universal operations}. Suppose

\begin{itemize} 

\item The promise $\pi_a$ equals:

\beq
\begin{array}{c}
P(z, y, x)
=
P(z|y)P(y|z)P(z)
\\
\\
\xymatrix{
\rvz
&\rvy\ar[l]
&\rvx\ar[l]}
\end{array}
\eeq

\item the promise $\pi_b$ equals:

\beq
\begin{array}{c}
Q(z, y, x)=Q(z|y, x) Q(y| x)Q(x)
\\
\\ 
\xymatrix{
\rvz
&\rvy\ar[l]
&\rvx\ar[l]
\ar@/_1pc/[ll]
}
\end{array}
\eeq
\end{itemize}

Note that 

\beq
\begin{array}{l}
\sum_y P(z|y)P(y|x)= P(z|x)
\\
\\
\xymatrix{
z
&\sum y\ar[l]
&x\ar[l]}=
\xymatrix{z&x\ar[l]}
\end{array}
\label{eq-markovian-op-only}
\eeq
and

\beq
\begin{array}{l}
\sum_y Q(z|y, x)Q(y|x)= Q(z|x)
\\
\\
\xymatrix{
z
&\sum y\ar[l]
&x\ar[l]
\ar@/_1pc/[ll]
}=
\xymatrix{z&x\ar[l]}
\end{array}
\label{eq-universal-op}
\eeq
An alternative notation is to
condition on the promise:

\beq
P(\cdot)= P(\cdot\mid \pi_a),
\quad Q(\cdot)=P(\cdot\mid \pi_b)
\eeq

Note that operation Eq.(\ref{eq-markovian-op-only})
is valid only for a Markovian
promise whereas operation
Eq.(\ref{eq-universal-op})
is a {\bf universal operation}, in the sense that it is valid for any promise.

Do Calculus can succumb to the
following two pifalls:

\begin{enumerate}
\item Making the
promise being considered  ambiguous 
 by neither writing
the conditioning on the promise,
nor using separate symbols $P(\cdot)$ and $Q(\cdot)$.
This does not cause problems
as long as we are dealing
with just one promise.
However, in the case
of Do Calculus,
we are dealing with
at least two
distinct promises,
the original promise (OP),
and its amputated version.
\item Using non-universal operations
to simplify an instantiation.
This removes memory of
how the instantiation was built (modulus universal operations which are not recorded). 
\end{enumerate}
A way to 
avoid these 2 pitfalls is to 
represent instantiations of
a bnet with diagrams, and
to use only 
 universal operations
to simplify a diagram.
Next, we shall discuss
universal operations
in more detail.

Consider the following  node summation identities.

\begin{enumerate}
\item

\beq
\sum_a
P(a|x_1, x_2)=1
\eeq

\beq
\xymatrix@R=.3pc{
x_1\ar[dr]
\\
&\sum a&=1
\\
x_2\ar[ur]
}
\label{eq-collider-sum}
\eeq
\hrule
\item
\beq
\sum_a P(y|a, x_1, x_2)P(a|x_1, x_2)P(x_2|x_1)P(x_1)
= P(y,x_1, x_2)
\;.
\eeq

\beq
\xymatrix{\\P(x_1)}
\xymatrix{
x_1\ar[rd]\ar[drr]\ar[dd]
\\
&\sum a\ar[r]
&y
\\
x_2\ar[ru]\ar[rru]
}
\xymatrix
{\\ =P(x_1)}
\xymatrix{
x_1\ar[rrd]\ar[dd]
\\
&&y
\\
x_2\ar[urr]
}
\label{eq-univ-bdoor}
\;.
\eeq
This identity can be interpreted
geometrically by noting that the 
bnet on the left-hand side (LHS) is fully connected, 
so one can reverse the direction of the 
arrow from $\sum a \rarrow y$ without changing the numerical value of the LHS. Then the sum over $a$
yields 1 so the $\sum a$ vertex 
with its 3 incoming arrows can be deleted.

The following identity has a similar geometrical interpretation

\beq
\sum_a P(y|x_2, a, x_1)P(x_2|a, x_1)P(x_1|a)P(a)
= P(y,x_1, x_2)
\;.
\eeq

\beq
\xymatrix{
x_1\ar[drr]\ar[dd]
\\
&\EE a\ar[r]\ar[ul]\ar[dl]
&y
\\
x_2\ar[rru]
}
\xymatrix
{\\ = P(x_1)}
\xymatrix{
x_1\ar[rrd]\ar[dd]
\\
&&y
\\
x_2\ar[urr]
}
\label{eq-univ-bdoor-ee}
\;.
\eeq


\hrule
\item
\beq
\sum_a P(y|a, x)P(a|x)P(x)=P(x,y)
\eeq

\beq
\xymatrix@R=1pc{
\\P(x)}
\xymatrix@R=1pc{
&\sum a\ar[rd]
\\
x\ar[ru]\ar[rr]
&&y
&=&P(x)\quad
x\ar[r]
&y}
\label{eq-med-sum}
\eeq
Once again, the bnet on the LHS is fully
connected so the arrow $\sum a\rarrow y$ can be reversed and then the sum over $a$ yields 1.

The following identity has a similar geometrical interpretation

\beq
\sum_a P(y|x,a)P(x|a)P(a)=P(x,y)
\eeq

\beq
\xymatrix@R=1pc{
&\EE a\ar[rd]\ar[ld]
\\
x\ar[rr]
&&y
&=&P(x)\quad
x\ar[r]
&y}
\label{eq-med-sum-ee}
\eeq

\hrule

\item
\beq
\sum_a P(y|a)P(a)=P(y)
\eeq

\beq
\xymatrix{
\EE a\ar[r]
&y
&=&P(y)
}
\label{eq-diff-priors}
\eeq
Furthermore,

\beq
\sum_a P(a|y)P(y)=P(y)
\eeq

\beq
\xymatrix{
P(y)\quad
\sum a
&y\ar[l]
&=&P(y)
}
\label{eq-diff-priors-ee}
\eeq

\end{enumerate}
\hrule

Recall that a  {\bf clique} 
of a DAG $\calg$ is a fully
connected subgraph of $\calg$.
Define an {\bf exploding clique} as one that
has no incoming arrows,
only outgoing ones.

Recall that a {\bf leaf node}
is a node with no outgoing arrows.

All of the above
node summation identities
are universal. They all
follow from  the
following 2  
{\bf basic universal
operations}:
\begin{mybluearea}
\begin{enumerate}
\item
Within an  exploding clique, one is free to
change the directions
of the arrows,
as  long as the 
final clique is still acyclic.
\item
Sums over leaf nodes
are always 1 
so they can be removed from or added to
the instantiation.
\end{enumerate}
\end{mybluearea}

%An important caveat concerning the above
%NS identities is that 
%the instantiations that they produce depend on the
%original bnet whence they come from. 
%This caveat is illustrated in Fig.\ref{fig-dep-on-origing-bnet}.
%This caveat is especial relevant when dealing 
%with do operators. In that case, the unamputated
%graph and the amputated graph are different bnets, with
%different promises.
%
%\begin{figure}[h!]
%$$
%\begin{array}{c|c}
%\sum_a P_{G}(x_2|a)P_{G}(a|x_1)=P_{G}(x_2|x_1)
%&
%\sum_a P_{G'}(x_2|a, x_1)P_{G'}(a|x_1)=P_{G'}(x_2|x_1)
%\\
%\xymatrix@C=1.5pc{
%x_1\ar[r]
%&\sum a\ar[r]
%&x_2
%&=&
%x_1\ar[r]
%&x_2}
%&
%\xymatrix@C=1.5pc{
%x_1\ar[r]\ar@/_1pc/[rr]
%&\sum a\ar[r]
%&x_2
%&=&
%x_1\ar[r]
%&x_2}
%\end{array}
%$$
%\caption{Note that there are cases when $P_G(x_2|x_1)
%\neq P_{G'}(x_2|x_1)$. Hence,
% $P(x_2|x_1)$ depends on the bnet of origin.
%}
%\label{fig-dep-on-origing-bnet}
%\end{figure}
%
%

\section{Not Do Calculus (NDC) Technique}


We define a {\bf do-adjustment formula}
for {\bf do-query}
$P(y|\cald\rvx=x)$
to be an algebraic expression  
constructed from the TPMs for {\it observed} (i.e., not hidden) nodes of the original bnet $G$.
If a do-adjustment formula
exists for a
particular do-query,
then we say the do-query is
 {\bf do-identifiable (DI)}.
% \footnote{
%To prove that a do-query $P(y|\cald\rvx=x, z)$
%is do-identifiable
%for a graph $G$,
%just prove that $\rvy\perp\rvx|\rvz$
%in $\call_\rvx G$.
%This is called Rule 2 of Do Calculus,
%but it is easy to understand just from
%the d-separation theorem.
%Info can be transmitted
%between $\rvy$ and $\rvx$ by
%either (1) paths
%in $\cald_\rvx G$
%or (2) paths in $\call_\rvx G$.
%$P(y|\cald\rvx=x, z)=
%P(y|x, z)$
%means the info is being transmitted only
%by (1).
%So the Rule 2 premise is checking
%that no info is being transmitted
%by (2). }
A {\bf do-transfer formula}
is a relationship between 2  do-queries.

Let $G$ be a graph before
amputation
of the arrows entering node $\rvx$,
and let $G_{amp(x)}=\cald_{\rvx=x}G$
be the same graph after amputation
at $\rvx$.
Also let $P_G(\cdot)$ be
the promise
for graph $G$, and $P_{G_{amp(x)}}(\cdot)$
be that for $G_{amp(x)}$.
In general, the
following is always true,
for bnets
with or bnets without hidden nodes:\footnote{Note that $P_{G_{amp(x)}}(y|x)\neq P_G(y|x)$
in general.
In fact,
$P_{G_{amp(x)}}(y|x)= P_G(y|x)$
iff there is no confounding,
so $P_{G_{amp(x)}}(y|x)\neq P_G(y|x)$
indicates confounding.
}

\beq
P_{G_{amp(x)}}(y|\cald\rvx=x)=P_{G_{amp(x)}}(y|x)\neq P_G(y|x)
\;
\eeq
However, $P_{G_{amp(x)}}(y|x)$ is
not a valid adjustment formula
for this query
because it's not expressed in terms
of observed TPMs of $G$.


%Note that
%it's always true that $P(y|\cald \rvx = x)=P(y|x)$.
%However, if the bnet contains more nodes than just $\rvx$ and $\rvy$, then the bnet is identifiable
%iff it can be stratified (i.e., 
%expressed as $P(y|x)=\sum_{z.}P(y|x, z.)P(z.)$
%for some observed multinode $\rvz.$).
%Thus, an alternative  name for identifiability is
%{\bf stratifiability}. Note also that there mightt be more than
%one adjustment formula, but they must all 
%be numerically equal to $P(y|x)$ for all $x$ and $y$.


The following 2 step heuristic technique,
which I call the {\bf Not Do Calculus} (NDC) technique, appears to be equivalent to Pearl's 3 rules of Do Calculus.\footnote{This NDC technique was invented solely by me. I haven't proven that it is equivalent to Pearl's 
3 rules of Do Calculus. I believe it is,
but it might not be. It 
correctly produces the backdoor and frontdoor adjustment formulae.}

\begin{mybluearea}

\begin{enumerate}
\item 
Equate $P(y\mid \cald\rvx=x)$
to a bnet instantiation
that has a DAG structure identical
to the OP,
except that arrows entering node
$\rvx$ have been amputated.
All nodes  of that instantiation, except
nodes  $x$ and $y$,
are summed over (indicated by $\sum$) if non-root nodes or averaged over
(indicated by $\EE$) if root nodes.



\item Use the basic universal operations to
remove all sums and averages over hidden nodes.
Each sum or average over a hidden node may be 
removed completely or replaced by a sum or average over an observed node. 
\end{enumerate}
Perform these steps until you remove all sums and averages  over hidden variables 
and obtain a valid adjustment formula.
If such an outcome cannot be achieved,
declare the  query unidentifiable.

\end{mybluearea}


\begin{figure}[h!]
$$
\begin{array}{ccc}
\xymatrix{
&\rvz\ar[ld]\ar[rd]
\\
\rvx\ar[rr]&&\rvy
}
&\quad\quad&
\xymatrix{
&*++[F-o]{\rvz}\ar[ld]\ar[rd]
\\
\rvx\ar[rr]&&\rvy
}
\\
(a) \text{$P(y|\cald\rvx=x)$ is DI}
&&(b) \text{$P(y|\cald\rvx=x)$ is not-DI}
\end{array}
$$
\caption{Examples of
OPs for which
the do-query $P(y|\cald\rvx=x)$
is
DI
and not-DI.
}
\label{fig-iden-noniden}
\end{figure}
See Fig.\ref{fig-iden-noniden}
for some simple
examples of OPs
 for which
the do-query $P(y|\cald\rvx=x)$
is
DI
and not-DI.
Using the NDC technique,
we can easily see why the
query $P(y|\cald\rvx=x)$
is DI
for OP $(a)$ 
and not-DI for OP $(b)$.

For OP $(a)$, after amputating arrow
$z\rarrow x$ and averaging over node $z$,
we get

\beqa
P(y|\cald\rvx=x)
&=&
\bcen
\xymatrix{
\EE z\ar[dr]
\\
x\ar[r]
&y
}\ecen
\label{eq-simplest-identifiable}
\eeqa
The right-hand
side of Eq.(\ref{eq-simplest-identifiable})
is a valid
adjustment formula
because it's expressed in terms
of observed TPMs of $G$.

For bnet $(b)$,
if we amputate arrow
$z\rarrow x$
and average over node $z$,
we get

\beqa
P(y|\cald \rvx =x)
&=&
\bcen
\xymatrix{
*++[F-o]{\EE z}\ar[rd]
\\
x\ar[r]
&y
}\ecen
\label{eq-simplest-not-DI}
\eeqa
The right-hand
side of Eq.(\ref{eq-simplest-not-DI})
is not a valid
adjustment formula
because
if we replace $\EE z$ by $\EE x' $, we get 
$\sum_{x'}P(y|x, x')P(x')$,
and $P(y|x, x')$ is undefined.
Furthermore,
there exists a
 $\delta_{hidden}$
such that

\beqa
\delta_{hidden}P(y|\cald \rvx =x)
&=&
\bcen
\xymatrix{
*++[F-o]{\delta\EE z}\ar[rd]
\\
x\ar[r]
&y
}\ecen
\neq 0
\eeqa


\section{Compilation of Do Calculus Proofs}
This section
is a compilation 
of proofs for
all the claims
in this book
that require Do Calculus to prove them. This includes
Pearl's Backdoor
and FrontDoor adjustment formulae,
and his Napkin problem.



\begin{claim}
\label{cl-decBackDoor}
\decBackDoor
\end{claim}

\proof

{\bf * proof 1:}
%\beq
%P(y|\cald\rvx=x, z)
%=
%\xymatrix{
%z\ar[dr]
%\\
%x\ar[r]
%&y
%}
%\eeq

\beq
P(y|\cald\rvx=x)
=
\bcen
\xymatrix{
\EE z\ar[dr]
\\
x\ar[r]
&y
}\ecen
\text{(by definition)}
\eeq


{\bf * proof 2:}
\begin{longtable}{l}
\color{red}
$P(y|\cald\rvx=x)
=
\sum_z
P(y|\cald\rvx=x, z)
P(z|\cald\rvx=x)$
\\
\quad by Probability Axioms
\\
\color{red}
$=\sum_z
P(y|x, z)
P(z|\cald\rvx=x)$
\\
\quad $P(y|\cald \rvx=x, z)\rarrow
P(y|x, z)$
\\
\quad  by Rule 2:
\ruletwoshort
\\
\quad
$\rvy\perp\rvx|\rvz$ in
$\call_\rvx\cald_\emptyset G:
\xymatrix{
\rvz\ar[d]\ar[rd]
\\
\rvx
&\rvy
}$
\\
\color{red}
$=\sum_z
P(y|x, z)
P(z)$
\\
\quad $P(z|\cald \rvx=x)\rarrow
P(z)$
\\
\quad  by Rule 3:
\rulethreeshort
\\
\quad
$\rvz\perp \rvx$ in
$\cald_\rvx \cald_\emptyset G:
\xymatrix{
\rvz\ar[rd]
\\
\rvx\ar[r]
&\rvy
}
$
\end{longtable}
\qed




\begin{claim}
\label{cl-decFrontDoor}
\decFrontDoor
\end{claim}

\proof


{\bf * proof 1:}

\begin{align}
P(y\mid \cald\rvx=x)&=
\xymatrix{
&*++[F-o]{\EE c}\ar[rd]
\\
x\ar[r]
&\sum m\ar[r]&y
}
\\
&=
\sum_c
\frac{1}{P(c|x)}
\bcen
\xymatrix{
&*++[F-o]{c}\ar@[red][rd]
\ar@[red][r]
& \sum x'
\\
x\ar[r]\ar[rru]\ar[ur]
&\sum m\ar[r]
&y\ar@[red][u]
}
\ecen
\begin{array}{l}
\text{(added summed leaf node $x'$,}
\\
\text{added $\rvx\to \rvc$ arrow)}
\end{array}
\\
&=\sum_c
\frac{1}{P(c|x)}
\bcen
\xymatrix{
&*++[F-o]{ c}\ar@{<-}@[red][rd]
\ar@{<-}@[red][r]
& \sum x'
\\
x\ar[r]\ar[rru]\ar[ur]
&\sum m\ar[r]
&y\ar@[red]@{<-}[u]
}
\ecen
\begin{array}{l}
\text{(fully connected 4 node}
\\
\text{graph so we are allowed} 
\\
\text{to reverse red arrows)}
\end{array}
\\
&=
\bcen
\xymatrix{
&*++[F-o]{\sum c}\ar@{<-}@[red][rd]
\ar@{<-}@[red][r]
& \sum x'
\\
x\ar[r]\ar[rru]
&\sum m\ar[r]
&y\ar@{<-}@[red][u]
}
\ecen
\text{(temporary arrow $\rvx\to \rvc$ removed)}
\label{eq-temp-ar-rem}
\\
&=
\bcen
\xymatrix{
& &\sum x'\ar[d]
\\
x\ar[r]\ar[rru]
&\sum m\ar[r]
&y
}
\ecen
\text{(removed summed leaf node $c$)}
\\
&=
\bcen
\xymatrix{
& &\EE x'\ar[d]
\\
x\ar[r]
&\sum m\ar[r]
&y
}
\ecen
\text{(used $P(x'|x)=P(x')$)}
\end{align}

The above argument seems to
be valid even if there is no node $\rvm$ in the promise.
However, with node $\rvm$,
the final adjustment
formula uses a $P(y\mid m, x)$
which is computable from the
available promise.
Without node $\rvm$,
we would need a $P(y\mid x, x')$
which is not computable from the
available promise.

Note that
we can prove that
$\delta_{hidden}P(y\mid \rvx=x)=0$,
but the differential
approach does not
yield an adjustment formula. Indeed,
by Eq.(\ref{eq-temp-ar-rem}), 
\beqa
\delta_{hidden}P(y\mid \rvx=x)&=&
\sum_c \delta c \pder{}{c}
\bcen
\xymatrix{
&*++[F-o]{ c}\ar@{<-}@[red][rd]
\ar@{<-}@[red][r]
& \sum x'
\\
x\ar[r]\ar[rru]
&\sum m\ar[r]
&y\ar@{<-}@[red][u]
}
\ecen
\\
&=& 0
\eeqa

%
%\begin{longtable}{l}
%$P(y|\cald\rvx=x) =\sum_c\sum_m P(y|m,c)P(m|x)P(c)$
%\\
%\hspace{2cm}
%\xymatrix{
%&\EE c\ar[dr]
%\\
%x\ar[r]&\sum m\ar[r]
%&y
%}
%\\
%\quad$=\sum_m \left[\sum_c\frac{1}{P(m|c)} P(y|m,c)P(m|c)P(c)\right]P(m|x)$
%\\
%\hspace{2cm}
%\xymatrix{\\
%\sum_m\sum_c\frac{P(c)}{P(m|c)}
%}
%\xymatrix{
%&&c\ar[dr]\ar[d]
%\\
%x\ar[r]&m&m\ar[r]
%&y
%}
%\\
%\quad$=\sum_m \frac{1}{P(m)}\left[\sum_c P(y|m,c)P(m|c)P(c)\right]P(m|x)$ \\
%\hspace{2cm}(because $P(m|c)=P(m)$ by d-separation
%on $G_{promise}|x$)
%\\
%\hspace{2cm}
%\xymatrix{\\
%\sum_m\frac{1}{P(m)}
%}
%\xymatrix{
%&&\EE c\ar[dr]\ar[d]
%\\
%x\ar[r]&m&m\ar[r]
%&y
%}
%\\
%\quad$=\sum_m P(y|m)P(m|x)$ 
%\\
%\hspace{2cm}
%(because square bracketed expression equals $P(y,m)$)
%\\
%\hspace{2cm}
%\xymatrix{
%x\ar[r]&\sum m \ar[r]& y
%}\quad (this simplifies frontdoor AF)
%\\
%\quad
%$=\sum_{x'}\sum_m P(y|m,x')P(m|x)P(x')$
%\\
%\hspace{2cm}
%(because can reverse steps that led to last expression)
%\\
%\hspace{2cm}
%\xymatrix{
%&\EE x'\ar[dr]
%\\
%x\ar[r]&\sum m\ar[r]
%&y
%}
%\end{longtable}
%
%\beqa
%P(y\mid \cald\rvx=x)
%&=&
%\bcen\xymatrix{
%&*++[F-o]{c}\ar[ld]\ar[rd]
%\ar[d]
%\\
%x\ar[r]\ar@/_1pc/[rr]
%&m\ar[r]&y
%}
%\ecen
%\eeqa
%Now,
%on both sides
%of the last equation,
% amputate arrows entering 
%node $\rvx$, 
%and sum over nodes that aren't
%$\rvx$ or $\rvy$
%
%\beq
%P(\rvy=y\mid \cald\rvx=x)=
%\bcen\xymatrix{
%&
%&\EE x'\ar[ld]\ar[d]
%\\
%x\ar[r]\ar@/_1pc/[rr]
%&\sum m\ar[r]
%&y
%}
%\ecen
%\eeq
%

{\bf * proof 2:}
\begin{longtable}{l}
$\color{red}
P(y|\cald\rvx=x)=
\sum_m
P(y|\cald\rvx=x, m)
P(m|\cald\rvx=x)$
\\
\quad by Probability Axioms
\\
$\color{red}=
\sum_m
P(y|\cald\rvx=x, \cald\rvm=m)
P(m|\cald\rvx=x)$
\\
\quad $P(y|\cald\rvx=x, m)\rarrow
P(y|\cald\rvx=x, \cald m=m)$
\\
\quad by Rule 2:
\ruletwoshort
\\
\quad $\rvy\perp \rvm|\rvx$ in
$\call_\rvm\cald_\rvx G:$
$\xymatrix{
&*++[F-o]{\rvc}\ar[rd]
\\
\rvx\ar[r]
&\rvm
&\rvy
}$
\\
$\color{red}=
\sum_m
P(y|\cald\rvx=x, \cald\rvm=m)
P(m| x)$
\\
\quad $P(m|\cald\rvx=x)\rarrow P(m|x)$
\\
\quad by Rule 2:
\ruletwoshort
\\
\quad
$\rvm\perp \rvx$ in
$\call_\rvx\cald_\emptyset G:$
$\xymatrix{
&*++[F-o]{\rvc}\ar[ld]\ar[rd]
\\
\rvx
&\rvm\ar[r]
&\rvy
}$
\\
$\color{red}=
\sum_m
P(y|\cald\rvm=m)
P(m|x)$
\\
\quad $P(y|\cald\rvx=x, \cald\rvm=m)
\rarrow
 P(y|\cald\rvm=m)$
\\
\quad by Rule 3:
\rulethreeshort
\\
\quad
$\rvy\perp\rvx|\rvm$ in
$\cald_\rvx\cald_\rvm G:$
$\xymatrix{
&*++[F-o]{\rvc}\ar[rd]
\\
\rvx
&\rvm\ar[r]
&\rvy
}$
\\
$\color{red}=
\sum_{x'}
\sum_m
P(y|\cald\rvm=m, x')
P(x'|\cald\rvm=m)
P(m|x)$
\\
\quad by Probability Axioms
\\
$\color{red}=
\sum_{x'}
\sum_m
P(y|m, x')
P(x'|\cald\rvm=m)
P(m|x)$
\\
\quad $P(y|\cald\rvm=m, x')
\rarrow
\quad P(y|m, x')$
\\
\quad by Rule 2:
\ruletwoshort
\\
\quad
$\rvy\perp\rvm|\rvx$ in
$\call_\rvm \cald_\emptyset G:$
$\xymatrix{
&*++[F-o]{\rvc}\ar[rd]\ar[ld]
\\
\rvx\ar[r]
&\rvm
&\rvy
}$
\\
$\color{red}=
\sum_{x'}
\sum_m
P(y|m, x')
P(x')
P(m|x)$
\\
\quad $P(x'|\cald\rvm=m)
\rarrow
P(x')$
\\
\quad by Rule 3:
\rulethreeshort
\\
\quad
$\rvx\perp\rvm$ in
$\cald_\rvm \cald_\emptyset G:$
$\xymatrix{
&*++[F-o]{\rvc}\ar[rd]\ar[ld]
\\
\rvx
&\rvm\ar[r]
&\rvy
}$
\end{longtable}
\qed



\begin{claim}
\label{cl-decNapkin}
\decNapkin
\end{claim}
\proof
\\
\begin{align}
P(y|\cald\rvx=x)
&=
\bcen
\xymatrix@R=1pc@C=1pc{
&*++[F-o]{\EE u_1}\ar[ddl]\ar[ddrr]
&&
\\
&*++[F-o]{\EE u_2}\ar[dl]
&&
\\
\sum w\ar[r]
&\sum z
&x\ar[r]
&y
}\ecen
\\
&=
\bcen \xymatrix{
*++[F-o]{\EE u_1}\ar[dr]
\\
x \ar[r]&  y}
\ecen
\end{align}
The last diagram is not identifiable
because  it's easy to find
a variation $\delta P(u_1)$
such that

\beq
\delta P(y|\cald\rvx=x) =
\bcen \xymatrix{
*++[F-o]{\delta \EE u_1}\ar[dr]
\\
x \ar[r]&  y}
\ecen \neq 0
\eeq



\qed


\begin{claim}
\label{cl-decWhy}
\decWhy
\end{claim}
\proof
\beqa
P(y|\cald\rvz=z, x)
&=&
\xymatrix{
x\ar[d]\ar[drr]
\\
 \sum w
&z\ar[r]
&y
\\
*++[F-o]{\EE u}\ar[u]\ar[urr]
}
\\
&=&
\xymatrix{
x\ar[drr]
\\
&z\ar[r]
&y
\\
*++[F-o]{\EE u}\ar[urr]
}
\\
&=&
\xymatrix{
x\ar[rrd]
\\
&z\ar[r]
&y
}
\eeqa
The last line follows because $P(u|x,z)=P(u)$ in 
$G_{promise}|(x,z)$ so
\beqa
\sum_u P(y|u,x,z)P(u)&=&
\sum_u P(y|u,x,z) P(u|x, z)
\\
&=&
P(y|x,z)
\eeqa
\qed

\begin{claim}
\label{cl-decTransferTrivial}
\decTransferTrivial
\end{claim}
\proof
\begin{align}
P^*(y|\cald \rvx=x, z) &=
P(y|\cald\rvx=x, z, \rvs=1)
\\
&=
\left[\frac{1}{\sum_{u_1}P(z|u_1)P(u_1)}\right]
\xymatrix@R=.5pc{
\EE u_1\ar@{-->}[r]
&z\ar[rd]
&s=1\ar[d]
\\
x\ar[rr]
&&y
\\
&\EE u_2\ar@{-->}[ul]\ar@{-->}[ur]
}
\\
& \text{(the factor in square brackets is necessary for normalization)}\nonumber
\\
&=
\xymatrix@R=.5pc{
&z\ar[rd]
&s=1\ar[d]
\\
x\ar[rr]
&&y
\\
&\EE u_2\ar@{-->}[ul]\ar@{-->}[ur]
}
\\
&=
\xymatrix{
&z\ar[rd]
&s=1\ar[d]
\\
x\ar[rr]
&&y}
\\
&=
P^*(y|x,z)
\end{align}
Let $\xi=(z,s)$. Above, we used 
\begin{align}
\sum_{u_2}P(y|x, u_2,\xi)P(x|u_2)P(u_2|\xi)
&=
\sum_{u_2}P(y|x, u_2,\xi)P(x|u_2, \xi)P(u_2|\xi)&=
\\&=
P(y|x, \xi)
\;\text{(because $P(x|u_2, \xi)=P(x|u_2)$ in $G_{promise}|\xi$)}
\end{align}

\qed

\begin{claim}
\label{cl-decTransferDirect}
\decTransferDirect
\end{claim}
\proof
\begin{align}
P^*(y|\cald x=x,z) &=
P(y|\cald x=x,z, \rvs=1)
\\
&=
\frac{1}{\sum_{u_1}P(z|s=1, u_1)P(u_1)}
\xymatrix@R=.2pc{
\EE u_1\ar@{-->}[dr]
\\
s=1\ar[r]
&z\ar[rd]
&
\\
x\ar[rr]
&&y
\\
&\EE u_2\ar@{-->}[ur]\ar@{-->}[ul]
}\\
& \text{(the factor in square brackets is necessary for normalization)}\nonumber
\\
&=
\xymatrix@R=.2pc{
&z\ar[rd]
&
\\
x\ar[rr]
&&y
\\
&\EE u_2\ar@{-->}[ur]\ar@{-->}[ul]
}
\\
&=
\xymatrix{
&z\ar[rd]
&
\\
x\ar[rr]
&&y
}
\end{align}
\qed

\begin{claim}
\label{cl-decTransferBox}
\decTransferBox
\end{claim}
\proof
\begin{align}
P^*(y|\cald x=x, a) &=
P(y|\cald x=x,a,\rvs=1)
\\
&=
\left[\frac {1}{P(a|z)P(z|s=1)P(z|u_1)P(u_1)}\right]
\xymatrix@R=.5pc{
\EE u_1\ar@{-->}[dr]
\\
s=1\ar[r]
&*++[F-o]{\sum z}\ar[r]
&a \ar[d]
\\
x\ar[rr]
&&y
\\
&\EE u_2\ar@{-->}[ur]\ar@{-->}[ul]
}
\\
&=
\xymatrix@R=.75pc{
&&a\ar[d]
&
\\
x\ar[rr]
&&y
\\
&\EE u_2\ar@{-->}[ur]\ar@{-->}[ul]
}
\\
&=
\xymatrix{
&&a\ar[d]
&
\\
x\ar[rr]
&&y
}
\end{align}
Finally, multiply both sides times $P^*(a)$
and sum over $a$.
\qed

%\decTransferOne merged
%with decDirectTransfer

\begin{claim}
\label{cl-decTransferNon}
\decTransferNon
\end{claim}
\proof

Obvious
\qed


\begin{claim}.
\label{cl-decTransferTwo}
\decTransferTwo
\end{claim}
\proof
\beqa
P^*(y|\cald \rvx=x)
&=&P(y|\cald \rvx=x, \rvs=1)
\\
&& \nonumber
\\
&=&
\xymatrix@R=.5pc{
s=1\ar@/^1.5pc/[rrr]
&\EE u_1\ar@{-->}[r]
&*++[F-o]{\sum h}\ar[dr]\ar[r]
&\sum z
\\
&x\ar[rr]
&
&y\\\
&&\EE u_2\ar@{-->}[ur]\ar@{-->}[ul]
}
\\
&& \nonumber
\\
&=&
\xymatrix@R=.5pc{
&\EE u_1\ar@{-->}[r]
&*++[F-o]{\sum h}\ar[dr]
\\
&x\ar[rr]
&
&y\\\
&&\EE u_2\ar@{-->}[ur]\ar@{-->}[ul]
}\\
&=& \text{independent of $s$}
\eeqa
\qed
\begin{claim}
\label{cl-decTransferThree}
\decTransferThree
\end{claim}
\proof

\beqa
P^*(y|\cald \rvx=x)
&=&P(y|\cald \rvx=x, \rvs=1)
\\
&=&
\xymatrix@R=.5pc{
s=1\ar[drr]
&\EE u_1\ar@{-->}[r]
&*++[F-o]{\su h}\ar[dr]
\\
&x\ar[r]
&\sum z\ar[r]
&y\\
&&\EE u_2\ar@{-->}[ur]\ar@{-->}[ul]
}
\\
&=&
\xymatrix
{\\
\sum_z}
\xymatrix{
s=1\ar[dr]
\\
x\ar[r]
&z
}
\xymatrix@R=.5pc{
\EE u_1\ar@{-->}[r]
&*++[F-o]{\sum h}\ar[dr]
\\
x
&z\ar[r]
&y\\
&\EE u_2\ar@{-->}[ur]\ar@{-->}[ul]
}
\\
&=&
\sum_z
P(y|\cald \rvx=x, \cald \rvz=z)P^*(z|x)
\eeqa
\qed

\begin{claim}
\label{cl-decMediationSimple}
\decMediationSimple
\end{claim}
\proof
Obvious.
\qed

\begin{claim}
\label{cl-decMediationPlus}
\decMediationPlus
\end{claim}
\proof
\beqa
P(y|\cald\rvd=d,\cali\rvd=d')&=&
\xymatrix{
&&*++[F-o]{\EE u}\ar[ddr]
\ar[dl]\ar[d]
\\
d'\ar@/_1pc/[rr]
&\sum\xi\ar[rrd]\ar[r]
&\sum m\ar[rd]
\\
d\ar[rrr]
&&&y}
\\
&=&
\xymatrix{
d'\ar@/_1pc/[rr]
&\EE\xi\ar[rrd]\ar[r]
&\sum m\ar[rd]
\\
d\ar[rrr]
&&&y}
\eeqa
\qed

\begin{claim}
\label{cl-decSeqBackDoor}
\decSeqBackDoor
\end{claim}
\proof

Obvious, once one defines 
\beq
[\cald \rvx^3 = x^3]=
\quad
[\cald\rvx_0=x_0 ,
\cald\rvx_1=x_1,
\cald\rvx_2=x_2]
\eeq
\qed

%\begin{claim}
%\label{cl-decSBBackDoor}
%\decSBBackDoor
%\end{claim}
%\proof
%\beqa
%P(y|\cald\rvx =x)
%&=& P(y|\cald\rvx =x, \rvs=0)
%\\
%&=&
%\xymatrix{
%\rvs=0\ar[r]\ar@/^1pc/[rr]
%&\sum z^-\ar[d]\ar[r]
%&\sum z^+
%\\
%x\ar[rru]\ar[r]
%&y
%}
%\\
%&=&
%\xymatrix{
%\rvs=1\ar[r]
%&\sum z^-\ar[d]
%\\
%x\ar[r]
%&y
%}
%\\
%&=&
%\xymatrix{
%\rvs=1\ar[r]\ar@[red][rd]
%&\sum z^-\ar[d]
%\\
%x\ar[r]
%&y
%}
%\\
%\nonumber
%&&
%\text{(because $P(y|x, z^-, s)=
%P(y|x, z^-)$ in $G_{promise}|x$)}
%\\
%&=&
%\xymatrix{
%\rvs=1\ar[r]\ar@[red][rd]
%&\sum z^-\ar[d]
%\\
%x\ar[r]\ar@[red][u]\ar@[red][ur]
%&y
%}
%\\
%\nonumber
%&&
%\text{(because $P(s|x)=P(s)$
%and $P(z^-|x)=P(z^-)$
%in $G_{promise}|x$)}
%\\
%&=&
%\xymatrix{
%\rvs=1\ar@[red][rd]
%\\
%x\ar[r]\ar@[red][u]
%&y
%}
%\\
%&=&
%\xymatrix{
%x\ar[r]
%&y
%}
%\\
%\nonumber
%&&
%\text{(because $P(y|x, s)=P(y|x)$
%and $P(s|x)=P(s)$
%in $G_{promise}|x$)}
%\eeqa
%\qed
%
%
%%\input{do-calc-proofs/do-calc-proofs-limbo.tex}
